---
title: "Analysis of FitBit data"
author: "Mikołaj Sobora"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/ASUS/Documents/Google Data Analitics Career Certificate Course/capstone_project/fitbit_dataset")
```

## Case Study

This is my first case study, showcasing acquired skills in R Programming. For this purpose I will use public data that explores habits of smart device users' daily habits. 

DATASET:
FitBit Fitness Tracker Data (CC0: Public Domain, dataset made available through Mobius): This Kaggle data set
contains personal fitness tracker from thirty fitbit users. Thirty eligible Fitbit users consented to the submission of
personal tracker data, including minute-level output for physical activity, heart rate, and sleep monitoring. It includes
information about daily activity, steps, and heart rate that can be used to explore users’ habits.

### Prepare

* Setting up my environment.

I dowloaded the dataset from Kaggle in .csv format and saved it on my harddrive. Then in Rstudio I installed necessary packages to proceed with my task.

```{r packages, message=FALSE, results='hide', warning=FALSE}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("data.table", repos = "http://cran.us.r-project.org")
install.packages("rmarkdown", repos = "http://cran.us.r-project.org")
library(data.table)
library(readr)
library(dplyr)
library(tidyr)
library(skimr)
library(janitor)
library(lubridate)
library(ggplot2)
library(scales)
library(rmarkdown)
```


Then I set my working directory to the file in which I saved this dataset using setwd() function and proceeded to load the .csv files.

```{r}
hourly_calories <- read.csv("mturkfitbit_export_3.12.16-4.11.16/Fitabase Data 3.12.16-4.11.16/hourlyCalories_merged.csv")
hourly_intensities <- read.csv("mturkfitbit_export_3.12.16-4.11.16/Fitabase Data 3.12.16-4.11.16/hourlyIntensities_merged.csv")
hourly_steps <- read.csv("mturkfitbit_export_3.12.16-4.11.16/Fitabase Data 3.12.16-4.11.16/hourlySteps_merged.csv")
weight_log_info <- read.csv("mturkfitbit_export_3.12.16-4.11.16/Fitabase Data 3.12.16-4.11.16/weightLogInfo_merged.csv")
daily_activity <- read.csv("mturkfitbit_export_4.12.16-5.12.16/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
sleep_day <- read.csv("mturkfitbit_export_4.12.16-5.12.16/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
```



* Getting acquainted with the data.  

After uploading all files into RStudio I began to examine it. I created an object containing all files to make my work more time efficient.


```{r}
all_files <- c(hourly_calories, hourly_intensities, hourly_steps, weight_log_info, daily_activity, sleep_day)
```

First I used glimpse() to get some insights from the data.

```{r}
glimpse(all_files)
```

Then to get more details about each file I used head() function. 

```{r}
head(hourly_calories)
head(hourly_intensities)
head(hourly_steps)
head(weight_log_info)
head(daily_activity)
head(sleep_day)
```
```{r}
n_distinct(daily_activity$Id)
n_distinct(sleep_day$Id)

nrow(daily_activity)
nrow(sleep_day)
```

After examining the results here is what I found:

* all files have one column in common which is column "Id" therefore merging is tempting
* files describing hourly activities have the same amount of rows
* there are only 3 types of objects (characters, numeric or integers)
* columns are written in PascalCase which is not to my liking so I will have to change it to SnakeCase
* there were more participants in daily_activity dataset than in sleep dataset

### Process

* Merging files

```{r}
data_by_hours <- bind_cols(hourly_calories, hourly_intensities, hourly_steps)
```

* Examining the new dataset

```{r}
head(data_by_hours)
colnames(data_by_hours)
```

* Deleting duplicate rows

```{r}
data_by_hours <- subset(data_by_hours, select = c(-4,-5,-8,-9))
```

* Renaming columns

```{r renaming, results='hide'}
data_by_hours <- data_by_hours %>% rename(id=Id...1, activity_hour=ActivityHour...2, calories=Calories, total_intensity=TotalIntensity, avg_intensity=AverageIntensity, steps=StepTotal)

colnames(daily_activity)
daily_activity <- daily_activity %>% rename(id=Id, activity_date = ActivityDate, total_steps = TotalSteps, total_distance = TotalDistance, tracker_distance = TrackerDistance, logged_activities_distance = LoggedActivitiesDistance, very_active_distance = VeryActiveDistance, moderately_active_distance = ModeratelyActiveDistance, light_active_distance = LightActiveDistance, sedentary_active_distance = SedentaryActiveDistance, very_active_minutes = VeryActiveMinutes, fairly_active_minutes = FairlyActiveMinutes, lightly_active_minutes = LightlyActiveMinutes, sedentary_minutes = SedentaryMinutes, calories = Calories)

colnames(sleep_day)
sleep_day <- sleep_day %>% rename(id = Id, sleep_day = SleepDay, total_sleep_records = TotalSleepRecords, total_minutes_asleep = TotalMinutesAsleep, total_time_in_bed = TotalTimeInBed)
```

```{r}
head(data_by_hours)
head(daily_activity)
head(sleep_day)
```

* Changing DATE format

```{r}
data_by_hours$activity_hour <- format(as_datetime(data_by_hours$activity_hour, format = "%m/%d/%Y %I:%M:%S %p"), "%d/%m/%Y %I:%M:%S %p")

data_by_hours$activity_hour <- parse_date_time(data_by_hours$activity_hour, "%d/%m/%Y %I:%M:%S %p")

str(data_by_hours)
```

### Analysis

* Summary statistics and visuals

```{r}
daily_activity %>%
  select(total_steps,
         total_distance,
         sedentary_minutes,
         calories) %>%
  summary()
```

Here I made basic summary of daily_activity dataset. The pattern seems clear: The more you walk the more you spend time resting and burn calories. But is it accurate for all tested subjects? Next is a visualization created on two arguments: total_steps and calories:

```{r, echo=FALSE}
ggplot(data=daily_activity) + geom_point(aes(x=total_steps, y=calories)) + ggtitle("Plot of Steps by Calories") + xlab("Steps") + ylab("Calories")
```

This viz negates previous insights. Steps and calories are not directly proportional which suggest other factors that are relevant to burnt calories. The dataset needs further examination. I decided to play some more with this two values and created a visualization showing steps and calories burnt of each fitbit user from a whole month.

```{r}
calories <- hourly_calories %>% group_by(Id) %>% 
  summarise(sum_calories=sum(Calories),
            .groups = 'drop')

steps <- hourly_steps %>% group_by(Id) %>% 
  summarise(sum_steps=sum(StepTotal),
            .groups = 'drop')

calories_steps_corellation <- merge(calories, steps, by ="Id")
```
```{r, echo=FALSE}
mindate <- min(data_by_hours$activity_hour)
maxdate <- max(data_by_hours$activity_hour)
ggplot(data=calories_steps_corellation) + geom_point(aes(x=sum_steps, y=sum_calories, color = as.factor(Id)), show.legend = FALSE) + scale_x_continuous(labels = label_comma()) + labs(title="Taken steps and burnt calories in a month time period", caption = paste0("Date from: ", mindate, " to ", maxdate),
       x="Steps",
       y="Calories")
```

Moving on to another summary:
```{r}
sleep_day %>%
  select(total_sleep_records,
         total_minutes_asleep,
         total_time_in_bed) %>%
  summary()
```

Just like before, this summary tells me that the more one sleeps the more he stays in bed. 

```{r, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(data=sleep_day, aes(x=total_minutes_asleep, y=total_time_in_bed)) + geom_point() + geom_smooth(method=lm, se=FALSE, col='blue', size=2) + labs(title = "Time in bed vs sleep time", x="Minutes asleep", y="Minutes in bed")
```

This visualization proves previous insight true. The trend line is very close to the points of this scatterplot which implies a strong correlation between minutes asleep and time spent in bed.

## Summary

In this Case Study I showed what I have learned in R programming regarding Data Analysis. I dowloaded and loaded dataset from a public source into RStudio, used some functions from tidyverse library, cleaned and processed the data, did few aggregation functions, plotted clear and easy to understand charts and managed to deliver a couple of insights based on the story that data showed me. I am open to constructive criticism on how to improve my skills and what could have bedn done better for next time. I am always open to acquiring new knowledge :)
